---
title: Evaluate how instrument selection approaches perform in excluding invalid instruments
author: Gibran Hemani
---

```{r, echo=TRUE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
	library(knitr)
	library(dplyr)
	library(tidyr)
	library(ggplot2)
	library(gridExtra)
})
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE, cache.lazy=FALSE)
```

Wrangle data

```{r }
load("../results/sim3/agg-instrument_validity.rdata")
load("../results/sim3/agg-param.rdata")
load("../results/sim3/agg-plei_summary.rdata")

validity <- inner_join(instrument_validity, param, by="id")
validity$strategy <- paste(validity$selection, validity$measure)
validity$n <- validity$nidx
validity$n[validity$hypothesis == "y"] <- validity$nidy[validity$hypothesis == "y"]
validity$nbin <- cut(validity$n, breaks=10, labels=FALSE)
plei <- plei_summary %>% group_by(id) %>% summarise(nonpl=sum(nonpl * effd / sum(effd)))
validity$code <- paste0(validity$id, validity$hypothesis)
validity <- inner_join(validity, plei, by=c("code"="id"))
validity$pbin <- cut(validity$nonpl, breaks=5, labels=FALSE)

```

We have a total of `r length(unique(paste(validity$id, validity$hypothesis)))`

Here we calculate the proportion of the selected variants that are direct effects, influencing confounders, or reverse causal effects

```{r}

validity_s <- group_by(validity, strategy, nbin) %>%
	summarise(
		n=sum(counts), 
		direct=sum(counts[type=="valid"])/n,
		confounder=sum(counts[type=="confounder"])/n,
		reverse=sum(counts[type=="reverse"])/n
	)
validity_sl <- gather(validity_s, key=key, value=value, direct, confounder, reverse)
validity_sl$nbin <- validity_sl$nbin * 50000
validity_sl <- filter(validity_sl, !grepl("^o", strategy))

p1 <- ggplot(filter(validity_sl, !grepl("^o", strategy)), aes(x=nbin, y=value)) +
geom_point(aes(colour=strategy)) +
geom_line(aes(colour=strategy)) +
facet_grid(. ~ key) +
scale_colour_brewer(type="qual") +
labs(x="Instrument discovery sample size", y="Proportion of instruments", colour="Selection", title="A") +
theme(legend.position=c(0.8, 0.7), legend.text=element_text(size=8), axis.text.x=element_blank(), axis.title.x=element_blank())

```

Here we get actual counts of the number of variants that are direct effects, influencing confounders, or reverse causal effects. This is averaged across the simulations.


```{r}
counts_s <- group_by(validity, strategy, nbin) %>%
	summarise(
		n=length(unique(paste(id, hypothesis))), 
		direct=sum(counts[type=="valid"])/n,
		confounder=sum(counts[type=="confounder"])/n,
		reverse=sum(counts[type=="reverse"])/n
	)
counts_sl <- gather(counts_s, key=key, value=value, direct, confounder, reverse)
counts_sl$nbin <- counts_sl$nbin * 50000
counts_sl <- filter(counts_sl, !grepl("^o", strategy))

p2 <- ggplot(filter(counts_sl, !grepl("^o", strategy)), aes(x=nbin, y=value)) +
geom_point(aes(colour=strategy)) +
geom_line(aes(colour=strategy)) +
facet_grid(. ~ key) +
scale_colour_brewer(type="qual") +
labs(x="Instrument discovery sample size", y="Number of retained instruments", colour="Selection", title="B") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5), legend.position="none")

```

```{r}
grid.arrange(p1, p2)
```

```{r}
temp <- inner_join(counts_sl, validity_sl, by=c("strategy", "nbin", "key"))

temp$key <- as.factor(temp$key)
temp$key <- factor(temp$key, levels=levels(temp$key)[c(2,1,3)])

ggplot(temp, aes(x=value.x, y=value.y)) +
geom_point(aes(colour=strategy, alpha=nbin)) +
geom_line(aes(colour=strategy)) +
facet_grid(key ~ ., scale="free_y") +
scale_colour_brewer(type="qual") +
labs(x="Number of retained instruments", y="Proportion of instruments", colour="Selection", alpha="Discovery\nsample size")

```

Stratify by pleiotropy level of the simulation. Higher values (1-5) indicate higher proportion of validity in the instruments

```{r}
counts_s <- group_by(validity, strategy, nbin, pbin) %>%
	summarise(
		n=length(unique(paste(id, hypothesis))), 
		direct=sum(counts[type=="valid"])/n,
		confounder=sum(counts[type=="confounder"])/n,
		reverse=sum(counts[type=="reverse"])/n
	)
counts_sl <- gather(counts_s, key=key, value=value, direct, confounder, reverse)
counts_sl$nbin <- counts_sl$nbin * 50000
ggplot(filter(counts_sl, !grepl("^o", strategy), !grepl("nofilter", strategy), !is.na(pbin)), aes(x=nbin, y=value)) +
geom_point(aes(colour=key)) +
geom_line(aes(colour=key)) +
facet_grid(pbin ~ strategy) +
scale_colour_brewer(type="qual") +
labs(x="Instrument discovery sample size", y="Proportion of instruments", colour="SNP\nassociation") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```


